---
title: "R4DS - Exploratory Data Analysis"
author: "Simon Thornewill von Essen"
date: "3/7/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```

## What is EDA

EDA is the cycle of asking questions about the data, answering these questions through data analysis and then using
what we learned to generate new questions. As we continue to go through this cycle, we hopefully start to understand
our data on a deeper level.

Note that this is a creative process that depends a lot of skill and experience as well as asking the right questions
instead of having a lot of "low quality" questions.

## Variation

As an example, the first questions we tend to ask about our data are ones of variation. "How does x change with respect
to y?" 

Variables can be categorical, such as we see below...

```{r}
ggplot(data=diamonds) + 
  geom_bar(mapping = aes(x=cut))
```

Note that we can get this bar chart in a table by using dplyr:

```{r}
diamonds %>%
  count(cut)
```

We can also get continuous variables

```{r}
ggplot(data = diamonds) +
  geom_histogram(mapping = aes(x = carat), binwidth = 0.5)
```

So, how would we generate this histogram in tabular format? The `count` function allows us to `cut_width` to create
bins which we then use for counting

```{r}
diamonds %>% 
  count(cut_width(carat, 0.5))
```

Note that setting the binwidth is something that is usually done through trial and error and can be difficult to set
algorithmically. 

