---
title: "R4DS - Data Vis"
author: "Simon Thornewill von Essen"
date: "2/11/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library("tidyverse")
```

In this notebook, I will write some notes about how to make data visualisations in R using the miles per gallon dataset.

We can find this dataset by typing `mpg` after importing the `ggplot2` package.

```{r}
mpg
```

To create a plot call `ggplot`, pass the dataframe that will be used and then map things accordingly after specifying 
the geom envrionment.

```{r}
ggplot(data=mpg) + 
  geom_point(mapping = aes(x=displ, y= hwy))
```

If you run ggplot on it's own you get an empty space.

```{r}
ggplot(data=mpg)
```

Not very interesting...

We can get the number of column and rows of a dataset by using the nrow and ncol functions.

```{r}
ncol(mpg)
nrow(mpg)
```

So, using the logic from above, we can get a scatter plot of the `hwy` and `cyl` variables like so.

```{r}
ggplot(data=mpg) + 
  geom_point(mapping = aes(x=hwy, y=cyl))
```

```{r}
ggplot(data=mpg) + 
  geom_point(mapping = aes(x=class, y=drv))
```

Obviously, this plot isn't very useful because we are looking at two categorical variables. The points are just telling
us that these observations exist, but tells us nothing more about these points.

We can color the points using the aesthetic function inside of the geoms:

```{r}
ggplot(data=mpg) + 
  geom_point(mapping=aes(x=displ, y=hwy, color=class))
```

We can how see where the different classifications of cars are concentrated in this graph. The colors are all mapped
automatically for us which saves us having to write something like a for-loop like we might for other visualisation
packages.

Note that we can use other mappings such as `alpha` and `shape` parameters

```{r}
ggplot(data=mpg) + 
  geom_point(mapping=aes(x=displ, y=hwy, alpha=class))
```

Luckily, we get a warning that using a categorical variable inside of alpha is likely an idea that doesn't make much
sense.

```{r}
ggplot(data=mpg) + 
  geom_point(mapping=aes(x=displ, y=hwy, shape=class))
```

We can see here that we only have 6 shapes that we can choose from, so using the shape parameter here isn't so great
either.

Note as well that continuous variables cannot be mapped using shapes.

Ofcourse, settings can also be chosen manually by placing them outside of the aesthetics function

```{r}
ggplot(data=mpg) + 
  geom_point(mapping=aes(x=displ, y=hwy), colour="blue")
```

Note that if you try and define the color inside of the aesthetics function, it will interpret it as being one class and
give it the red color. You can wrap the string in an `I()` function, but its probably just best to assign these things
outside of the `aes` in the first place.

## Facets

Sometimes you want to plot multiple variables at a time, facets are good for this.

```{r}
ggplot(data=mpg) + 
  geom_point(mapping=aes(x=displ, y=hwy)) + 
  facet_wrap(~class, nrow=2)
```

Here, we were able to take the above graph and split it apart by the car class, which can sometimes be handy.

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(drv ~ cyl)
```

Here's a question, what happens if you facet on a continuous variable

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid( ~ cty)
```

It seems like it kind of works.

Note that faceting and using aesthetics have different functions. You might use faceting if you want to have everything
in one place to make comparisons between things more apparent, faceting separates the effects out so you can investigate
more inside of a single plot without having to faff around with filtering things.

## Geometric Objects

Note that the geom associated with a plot decides how the plot is going to be formed. 

```{r}
# left
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))

# right
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy))
```

Here, we get two plots, but one is some kind of a model that helps us to see the underlying trend in the data, we can
also combine these plots together as follows.

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))+   # left
  geom_smooth(mapping = aes(x = displ, y = hwy))  # right
```

It seems like this model is overfitting to the data a little bit. The couple of points above seem to be drawing the line
further up than would probably make sense.

Note that we can also map the smoothing function to things like what kind of a drive the car is.

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +  
  geom_smooth(mapping = aes(x = displ, y = hwy, color=drv, linetype=drv))   
```

Note that you can also create separate lines using the `group` aesthetic. It'll perform the same function as when you 
use the color aesthetic, only the lines won't be colored. Shocker!

This line is overfit even more because the model that is used to fit the data is very flexible but we dont have enough
points to justify using something so flexible, except maybe for four wheel drive cars, there the model seems to fit a
little better.

If you ever want to hide a legend, you can pass a `show.legend` parameter

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +  
  geom_smooth(mapping = aes(x = displ, y = hwy, color=drv, linetype=drv), show.legend = FALSE) 
```

That's quite useful to know!

Note that above we have some duplication when we combined the points geom with the smooth geom, what if there were a way
to program this all together...

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()
```

Anything that's defined in the aesthetics section of the parent `ggplot` will be passed to all of the geoms as well.

You can even extend what was set by default by further specifying things inside of the geoms.

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(aes(color=class)) + 
  geom_smooth()
```

You can even specify which points to filter for.

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(aes(color=class)) + 
  geom_smooth(data=filter(mpg, class!="2seater"), se=FALSE)
```

Here, our smooth geom is ignoring the 2 seaters which were annoying from before, I think I might also ignore displ that 
are higher than 6.

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(aes(color=class)) + 
  geom_smooth(data=filter(mpg, class!="2seater", displ<6), se=FALSE)
```

Note that the `se` parameter dictates whether loess smoothing comes with a confidence interval.

We can also change it so that it does a linear regression.

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(aes(color=class)) + 
  geom_smooth(data=filter(mpg, class!="2seater", displ<6), method='lm')
```

## Statistical Transformations

Consider the diamonds dataset, we can use a barplot to get the count of each kind of cut

```{r}
ggplot(data=diamonds) + 
  geom_bar(mapping=aes(x=cut))
```

Something interesting just happened here, though! There is no "count" variable in the diamonds dataset, so what happened
here?

* Bar charts, histograms, and freq polygons automatically bin the data and plot the counts
* Smoothers fit a model to the data and then plot that fit
* boxplots compute a summary of the distribution and then plot that

Note that the statistic that is calculated can also be changed using the `stat` parameter.

You can create the same plot like so

```{r}
ggplot(data=diamonds) + 
  stat_count(mapping=aes(x=cut))
```

This works because the default stat for the bar plot is `count`!

Here's a difficulty though, what is the statistics have already been calculated?

```{r}
demo <- tribble(
  ~cut,         ~freq,
  "Fair",       1610,
  "Good",       4906,
  "Very Good",  12082,
  "Premium",    13791,
  "Ideal",      21551
)

demo
```

```{r}
ggplot(data=demo) + 
  geom_bar(mapping=aes(x=cut, y=freq), stat="identity")
```

Note that you need the `identity` stat because the stat that is being used is the value itself.

We can also do something like calculating the percent of total

```{r}
tot <- sum(demo$freq)
demo$perc.tot <- demo$freq/tot

ggplot(data=demo) + 
  geom_bar(mapping=aes(x=cut, y=perc.tot), stat="identity")
```

You can do a similar kind of thing using the diamonds dataset except without the explicit calculation.

```{r}
ggplot(data=diamonds) + 
  geom_bar(mapping=aes(x=cut, y=stat(prop), group=1))
```

I'm not exactly sure why we are using `group = 1` and why that gives the same thing when we equate it to two. I guess
it means that it treats all things inside of this vis as a group and then calculates the proporitons on that. If we
remove the group function then the proportion is always equal to one, as it groups by the cut.

It should also be noted that you can also get `geom_col` which requires that you pass whatever you are plotting on the 
y axis manually.

You can also do more sophisticated things

```{r}
ggplot(data=diamonds) + 
  stat_summary(
    mapping=aes(x=cut, y=depth),
    fun.ymin=min,
    fun.ymax=max,
    fun.y=median
  )
```

Note that ggplot has over 20 stats that can be used, each of which is a function. So you can use the `?stat_bin` 
function in order to have a look at all of them.

## Position Adjustments

You can also create barplots which are unstacked using colour vs fill.

```{r}
ggplot(data=diamonds) + 
  geom_bar(mapping = aes(x=cut, fill=cut))
```

```{r}
ggplot(data=diamonds) + 
  geom_bar(mapping = aes(x=cut, fill=clarity))
```
 
 The differnce here is the `position` parameter. This parameter can be filled with `identity`, `dodge` and `fill`.
 
```{r}
ggplot(data=diamonds) + 
  geom_bar(mapping = aes(x=cut, fill=clarity), position = "dodge")
```
 
```{r}
ggplot(data=diamonds) + 
  geom_bar(mapping = aes(x=cut, fill=clarity), position = "fill")
```

So, using this we can get different kinds of batplot, which is handy.

There is also a jitter position for scatterplots:

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), position = "jitter")
```

This can be useful to avoid overplotting in some cases, for instance.

## Coordinate Systems

You can also flip a graph by using the `coord_flip()` geom

```{r}
ggplot(data = mpg, mapping = aes(x = class, y = hwy)) + 
  geom_boxplot() +
  coord_flip()
```
