---
title: "R4DS - Data Tansformation"
author: "Simon Thornewill von Essen"
date: "3/7/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(nycflights13)
library(tidyverse)
```

Alright, so data visualisation can help us to understand our data relatively quickly. The only issue is that we rarely
get the data in the format that we need. Towards, this end, we need to be able to manipulate data within R.

The dataframe we'll be exploring is the `flights` dataframe

```{r}
head(flights)
```

Note that in addition to being a dataframe, this dataframe is also a tibble. 

Note that the columns also have data types:

* int $\leftarrow$ integer
* dbl $\leftarrow$ double
* chr $\leftarrow$ character/string
* dttm $\leftarrow$ date-time

There are also other variables that exist:

* lgl $\leftarrow$ boolean
* fctr $\leftarrow$ factor
* date $\leftarrow$ date

## dplyr basics

Here, we will use dplyr, which has the following useful functions:

* filter
* arrange
* select
* mutate
* summarise

All of these can be used in conjunction with `groupby`.

The logical workflow is like this:

1. You select the dataframe you want to work with
2. The subsequent arguments say what you want to do with it
3. You then get an output dataframe

## Filtering

Let's try and look at flgihts that take place on the first of jan

```{r}
filter(flights, month==1, day==1)
```

Can we also do it like this?

```{r}
flights %>% filter(month==1, day==1)
```

Yes! It also works like this. We can select a dataframe and then perform things on it using the `%>%` character.

Now, let's save this to a new variable.

```{r}
jan1 <- flights %>% filter(month==1, day==1)
```

Note that if you use an NA then that will be propogated, you cannot check if a variable is NA simply by typing
`var == NA` because the result will also be NA.

Thus, you need to use the is.na function;

```{r}
is.na(NA)
```

## Order rows with arrange

Next, if we want to view the flights from jan1 in some order, say we want to order by carrier, then we can use the 
arrange function.

```{r}
arrange(jan1, carrier)
```

We can also chain these ideas together like so.

```{r}
flights %>% 
  filter(month==1, day==1) %>% 
  arrange(carrier)
```

## Select with select

Afterwards, we can select relevant columns that we are interetsed in

```{r}
flights %>% 
  filter(month==1, day==1) %>% 
  arrange(carrier) %>%
  select(year, month, day)
```

Note that we can change the order as well

```{r}
flights %>% 
  arrange(carrier) %>%
  select(year, month, day) %>% 
  filter(month==1, day==1) 
```

Note that there are interesting helper functions we can use within select

* `starts_with` matches names that start with a string
* `ends_with` matches names that end with a string
* `contains`: matches names that contains a string
* `matches`: matches columns with a regular expression
* `num_range`: appends numbers in a sequence at at the end of a string and searches the columns for that

You can also reorder a dataframe by using `everything`, which will select the rest of the columns that are not 
mentioned.

As an example, you could do the following:

```{r}
flights %>%
  rename(date_year=year, date_month=month, date_day=day) %>%
  select(starts_with("date_"), everything())
```

## Mutate

Alright, so we know how to slect, order and filter, but what about if we want to make a new column

For this, we use the mutate function:

```{r}
flights %>%
  select(year:day,
         ends_with("delay"),
         distance,
         air_time) %>%
  mutate(gain = dep_delay - arr_delay,
         speed = distance / air_time * 60)
```

Now, we have two columns for gain and speed, which we got from mutating columns that we already had.

If we only want to keep the variables we mutate, we can also transmute them:

```{r}
flights %>%
  select(year:day,
         ends_with("delay"),
         distance,
         air_time) %>%
  transmute(gain = dep_delay - arr_delay,
         speed = distance / air_time * 60)
```

Note that the mutate function also accepts lag and lead functions, which can be very handy. It can also calculate
cumulative values as well was row numbers and this kind of thing.

## Summarise

Finally, after all of that, we can aggregate our data to get summaries.

```{r}
summarise(flights, delay = mean(dep_delay, na.rm = TRUE))
```

We can pair `summarise` with `groupby` which will select only certain columns to group on for aggregations.

Note that summarise allows you to use multiple aggregations (or even functions, it seems) as an input so that you dont
need to use only one aggregations like you might by grouping a pandas dataframe in python.

Some interesting functions to use are:

* mean
* median
* sd
* IQR
* mad (mean absolute deviation, I guess)
* min
* max
* quantile
* first
* nth
* last
* n (returns size of group)
