---
title: "R4DS - Tibbles"
author: "Simon Thornewill von Essen"
date: "9 3 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```

So far we have been working with tibbles, which are dataframes. Dataframes are an old datastructure and it can be difficult to change how they work without breaking other things. Tibbles are thus an updated version of a dataframe to suit modern needs.

You can take a dataframe and convert it to a tibble like this

```{r}
as_tibble(iris)
```

You can also create tibbles from individual vectors:

```{r}
tibble(x=1:5, y=1, z = x^2 + y)
```

There are a couple of things that a tibble does not do that a datafarme does, it doesnt convert inputs and it doesnt change the names of variables or create row names. It's also possible to have column names that are normally otherwise invalid.

tribbles, transposed tibbles, can also be created as follows:

```{r}
tribble(
  ~x, ~y, ~z,
  #--|--|----
  "a", 2, 3.6,
  "b", 1, 8.5
)
```

Notice that column names are created with tildes and each line represents a row.

The two main differences between Tibbles and dataframes have to do with how they print info and how you might pull information.

```{r}
df <- tibble(
  x = runif(5),
  y = rnorm(5)
)
```

You can call the columns like normal using the `$`.

```{r}
df$x
```

You can also call it like this:

```{r}
df[[1]]
```

or this

```{r}
df[, "x"]
```

You can also use the pipe as before, but you will need to use a `.` as a placeholder.

```{r}
df %>% .$x
```

