---
title: "R4DS - Data Import"
author: "Simon Thornewill von Essen"
date: "9 3 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```

In this chapter, we'll have a look at importing data.

There are multiple ways to read in data, but the most important way is through csv files, which are the most common filetype. (This is done with `read_csv()`)

Sometimes you might also get fixed width files (fwf), so you should also know that you can use `read_fwf()`.

If you've used base-R then you might wonder about the differences between `read.csv()` and `read_csv()`. I actually didnt realise that `read_csv()` is a readr function until now.

* The readr function is faster
* It produces a tibble
* Its more reproducible

## Parse Functions

Before going further, we should learn about `parse_*()` functions such as `parse_logical`, `parse_integer` and `parse_date` since they are important building blocks of readr.

The first argument is always the vector to be parsed, as well as an na parameters to search for missing values.

```{r}
parse_integer(c("1", "231", ".", "456"), na=".")
```

## Numbers

When it comes to parsing numbers, it can be a pain because different parts of the world switch around `.` and `,` and they can often be followed by units such as `$` or `%`which make them a pain to interpret.

```{r}
parse_double("1,23", locale = locale(decimal_mark = ","))
```

```{r}
parse_number("$100")
```

```{r}
parse_number("It cost $123.45")
```

Note that readr is able to handle all of these cases flawlessly.

The `read_character` function is also able to handle multiple languages and character that don't belong to the standard english.

```{r}
x2 <- "\x82\xb1\x82\xf1\x82\xc9\x82\xbf\x82\xcd"

parse_character(x2, locale = locale(encoding = "Shift-JIS"))
```

Note that readr can also guess the encoding of an input string, which is very handy.

```{r}
# guess_encoding(charToRaw(x2))

parse_character(x2, locale = locale(encoding = "KOI8-R"))
```

Didn't work particularly well in this case, but it was also not super confident about his choice.

Note that we can also parse factors, as well as dates, datetimes and times.

```{r}
parse_date("2019-01-01")
```

Note that if you use a different format, you can pass the format to the parser

```{r}
parse_date("2019/01/01", "%Y/%m/%d")
```

If you are parsing from a language, you can also specifcy the language

```{r}
parse_date("1 janvier 2015", "%d %B %Y", locale = locale("fr"))
```

Note that this locale variable is similar with how you might parse a localse with other parse function, which makes learning
all of this a little easier.

If readr has a hard time using its heuristics to figure out what column type a column should be, you can also pass the `col_types` variable with `cols(x = col_double(), y=col_date())` as an example.

## Writing to a file

Finally, if you cant to spit out a csv you can use the `write_csv()` function. You can even export to an excel file as well with `write_excel_csv()`. You can also write to rds, which is R's custom binary format, with `write_rds()`.

Note that you might need other functions as well in the case that you need to read:

* SPSS, Stata or SAS files $\rightarrow$ `haven`
* xlsx and xls files $\rightarrow$ `readxl`
* Relational databases $\rightarrow$ `DBI`