---
title: "R4DS - Tidy Data"
author: "Simon Thornewill von Essen"
date: "9 3 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```

One thing that the tidyverse does is expect the data to be in a certain format. This is powerful because if the data is in a format that can be consistently expected then lots of functions can be written around that.

This format is called "tidy" data. Data is tidy when each column represents a single variable and each row represents a single observation and each cell contains a single value.

Note that when two of these conditions are met, the third is implied. 

## Pivoting Data

The only issue is that our data won't always be in this representation, consider the following table.

```{r}
print(table4a)
```

We can see here that there are two columns for the year, but this then means that the data isn't tidy. How can we take these columns and stack them?

We can do this using the `pivot_longer` function to melt the columns.

```{r}
table4a_longer <- table4a %>%
  pivot_longer(c("1999", "2000"), names_to="year", values_to="cases")

print(table4a_longer)
```

It's really handy to know how to do something like this. We can then join on other tables to slowly build our dataset.

```{r}
tidy4b <- table4b %>% 
  pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "population")

table4a_longer %>%
  left_join(tidy4b)
```

Obviously, the opposite is to pivot the data using `pivot_wider`.

```{r}
table2
```


```{r}
table2 %>%
  pivot_wider(names_from = type, values_from = count)
```

What format you want your dataset in depends a little on your application. But this is the general gist of it.

## Separating and unifying columns

The issue with `table3` requires something more sophisticated to tidy the data

```{r}
table3
```

We can see here that the rate is a fraction of two numbers. We should create separate columns for these two because it gives us the most flexibility with what we can use for data for.

```{r}
table3 %>% 
  separate(rate, into = c("cases", "population"))
```

From here, we might want to consider melting these columns again.

There are multiple ways you might separate the columns, but I leave that for the documentation. `?separate`
